Experience Report: Haskell in Computational Biology

### 2012-03-09 mrfy_experience_report.tex
working on 'perspicuous haskell'

@@

The states of the hidden Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}.
The match state determines the probability that the query sequence
has a particular amino acid in column~$j$.
The match state assigns high probability to the amino acids that
appear in column~$j$, and a low but nonzero probability to other amino
acids.
A~match state is more probable if it corresponds to a consensus column.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
A~deletion state is more probable if it corresponds to a
non-consensus column.
% has high probability if there are lots of gaps in
%column~$j$, and low probability if there are few gaps in column~$j$.
%Thus, the deletion state is more likely in non-consensus columns.
\item
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}.
The insertion state determines the probability that a query sequence
has an ``extra'' amino acid.
Like a match state, the insertion state assigns high probability to
amino acids appearing in column~$j$.
An~insertion state is more probable if it corresponds to a
non-consensus column.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
The probability associated with a specific state is conditional upon the previous
state in the model, so it is a \textit{transition probability}.
Mathematically, transition probabilities are represented by $a_{s_{j-1}s_{j}}$
for a state $s_{j}$.
In Haskell, we use 
\begin{smallverbatim}
type logProb = Double
data TProb  = TProb { logProbability::LogProb
                    , fromState::HMMState
                    , toState::HMMState
                    }
\end{smallverbatim}
to represent the seven allowed state transitions.

We~write a query sequence as
$x_{1}, \ldots, x_{n}$, where each~$x_i$ is an amino acid.
In~idiomatic Haskell, we might represent each individual amino acid
using a value of algebraic data type, thus:
\begin{smallverbatim}
data Amino = Ala | Cys | Asp | Glu | Gly | ...
\end{smallverbatim}
But our primary goal was to get something working quickly with a
legacy file format that represents each amino acid as a small integer.
Our types are therefore
\begin{smallverbatim}
type AA       = Int -- amino acid using HMMER coding
type QuerySeq = Vector AA
\end{smallverbatim}

Thus, the hidden Markov model can be represented as a directed graph, with each
edge weight corresponding to the probability of a specific state transition.
\figref{plan7} shows a hidden Markov model derived from an alignment
with just four columns.
Each column in the original alignment corresponds to a \emph{node},
which is a triplet holding an insertion state, a deletion state, and a
match state.
Mathematically, nodes are simply $\left\{a_{s_{k},s_{j}}, e_{s_{j}}(x_{i})\right\}$
for states $s_{k}, s_{j}$ and emission probabilities $e_{s_{j}}(x_{i})$ for a
query sequence element $x_{i}$.
In Haskell, we represent a node as a group of transition probabilities
\begin{smallverbatim}
data TProbs = 
     TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
            , i_m :: TProb, i_i :: TProb, d_m :: TProb
            , d_d :: TProb}
\end{smallverbatim}
and a vector of emission probabilities for the match and insertion states.
\begin{smallverbatim}
type EProbs = V.Vector logProb
data HmmNode = HmmNode { nodeNum :: Int
                       , matEmissions :: EProbs
                       , insEmissions :: EProbs
                       , transitions :: TProbs
                       }
\end{smallverbatim}

>>>>

\newcommand\txprobj[3][]{a#1_{#2_{j-1}#3_j}}
\newcommand\txprobjj[3][]{a#1_{#2_{j-1}#3_j}}

@@

A~hidden Markov model carries probabilities on its states and on its
state transitions.
Both the probabilities and the
the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
This probability is called an \emph{emission} probability.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\end{itemize}
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.

The particular hidden Markov model we use requires that every nonempty
sequence of insertions be followed by a match, and similarly for
deletions.
\remark{citation needed?}
In~other words, insertions and deletions cannot be mixed without an
intervening match.
This restriction means that between two adjacent columns, six
transitions are possible.
An~insertion state may also transition to itself.
\begin{itemize} 
\item
A~transition into a match state in
column~$j$ is more likely when column~$j$ is a consensus column.
Depending on the predecessor state, 
the probability of such a transition is notated
$\txprobj M M$, $\txprobj I M$, or $\txprobj D M$.
\item
A~transition into a deletion state in
 column~$j$ is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is notated
$\txprobj M D$ or $\txprobj D D$.
\item
A~transition into an insertion state in
 column~$j$ is relatively unlikely,
but it is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is notated
$\txprobjj M I$ or $\txprobjj I I$.
\end{itemize}

\subsection{Perspicuous Haskell}

Computation of probabilities from hidden Markov models is well
understood and is implemented in standard software packages.
To~be able to use Haskell, we had to reimplement this standard
algorithm.
Haskell made it possible for us to write code that looks like the
math,
which made the code easy to write and gives us confidence that it is
correct.

We~begin with our representations.
In~idiomatic Haskell, a query sequence is an immutable vector, and
we might represent an individual amino acid~$x$
using a value of algebraic data type, thus:
\begin{smallverbatim}
data Amino = Ala | Cys | Asp | Glu | Gly | ...
\end{smallverbatim}
But our primary goal was to get something working quickly with a
legacy file format that represents each amino acid as a small integer.
Our representations are therefore
\begin{smallverbatim}
type AA       = Int -- amino acid using HMMER coding
type QuerySeq = Vector AA
\end{smallverbatim}

A~transition probability is simply a number.
For accuracy in computation, we add log probabilities instead of
multiplying probabilities.
And to distinguish a transition probability from other numbers, we wrap it in
the value constructor \texttt{TProb}:
\begin{smallverbatim}
type LogProb = Double
data TProb  = TProb { logProbability :: LogProb }
\end{smallverbatim}

We group transition probabilities into 7-tuples.
Each tuple corresponds to a ``node'' in the hidden Markov model
(a~dashed ovel in \figref{plan7}), and to a column of the original
alignment.
As~noted above, 
a match state may transition to any state in the successor node;
an insertion state may transition to itself or to its successor match
state;
and
a~deletion may transition to its successor match
state or deletion state.
\begin{smallverbatim}
data TProbs = 
     TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
            , i_m :: TProb, i_i :: TProb
            , d_m :: TProb, d_d :: TProb }
\end{smallverbatim}

The
and a vector of emission probabilities for the match and insertion states.
\begin{smallverbatim}
type EProbs = V.Vector LogProb
data HmmNode = HmmNode { nodeNum :: Int
                       , matEmissions :: EProbs
                       , insEmissions :: EProbs
                       , transitions :: TProbs
                       }
\end{smallverbatim}
